The tracing of trade

### 2021-02-10 Manuscript/Manuscript.tex
review hasta 3

The aim of this study is to explore the trade dynamics involved on the large-scale distribution of olive oil produced in \textit{Baetica} province (currently Andalusia). \textit{Baetica} became an important production and distribution centre during the Roman Empire and its goods have been found in thousands of sites, but it is still unclear how exactly this large scale production was organised.

We explore here connections between production and consumption sites during the period of most intense activity (1st to 3rd centuries AD). Specifically, we use amphoric stamps as proxies to identify the workshops where amphorae found on consumer sites were produced.

To achieve this goal, we analyse a dataset of stamps found in two types of sites: 1) workshops from \textit{Baetica} province and 2) consumption centres located in two provinces \textit{Germania} (Inferior and Superior) and \textit{Britannia}. This dataset is used to detect links between specific sites that may share a large percentage of common stamps. The analysis is performed using quantitative methods borrowed from Ecology that allows us to group sites based on the amount of stamps shared between them. Results highlight the structure of economic processes of the Roman Empire using the archaeological evidence of olive oil large-scale distribution.

@@

The Empire developed a series of structures to support and organise this long-range trade and specialised entire regions to produce specific goods at a massive scale. The province of \textit{Baetica} (currently Andalusia, southern Spain) is an example of this specialization process. \textit{Baetica} became one of the most important olive oil production centres during the Roman Empire; olive oil was an essential good for Romans because it was used in almost every aspect of their daily life such as cooking, hygiene or lighting~\citep{mattingly_d.j._oil_1988}. This high demand required a huge increase on the production which needed to be distributed to cities across thousands of kilometers. This large-scale trade was organised via the shipment of millions of amphorae using maritime and riverine routes to all the provinces and particularly to Italy and the military garrisons guarding the borders of the Empire~\citep{blazquez_exportacion_1980}.

The structure and processes associated with this massive olive oil production have been extensively discussed over the last decades\citep{rodriguez_economioleicola_1977, duncan1982economy, Chic_hispania_1997,millet_anforas_1998}. However, the complexity and sheer scale of the process makes its understanding challenging despite several archaeologicale xcavations and the existence of some written records discussing its structure. Our knowledge of this economic activity has benefited from new findings and data sources, but despite all these advances several questions remain open: How was this massive amphora production organised? Can we detect the economic ties between provinces? Was this distribution centralized or each region organised it differently?

Recent advances in Roman studies have currently led to an environment with various interpretations on these commercial dynamics~\citep{temin_economy_2006,quantifyingwilson2009}. The application of diverse novel quantitative approaches to new accessible datasets has allowed us to improve our interpretation of the growing amount of archaeological evidence and contribute to a better understanding of the dynamics of Roman production\citep{brughmans_roman_2016,orengo_seeds_2016,bayesian_2018,coto-sarmiento_identifying_2018,rubio-campillo_ecology_2018}. This work aligns with these efforts and focuses on the distribution of olive oil towards Roman provinces in the borders. Amphoric production in \textit{Baetica} province will help us identify links between production centres and consumption centres in order to detect an economic pattern in the commerce by analysing amphoric stamps finds in these provinces.

The finding of identical amphoric stamps in different provinces has allowed us to explore patterns of olive oil distribution within the Empire. Despite this archaeological evidence there is still no consensus on the structure that generated this pattern; how was the economic relation between production and consumption centres? Did they follow a centralised process to distribute olive oil in those provinces or each workshop organised its own trade networks?

This paper aims to explore these questions regarding the olive oil market by identifying links between productionn centres and consumption provinces using similarity indexes for amphoric stamps. In this way, our work aims to detect regional patterns for a specific commercial product (i.e. olive oil) which was shipped between distant provinces~\citep{isaksen_network_2006}.

@@

In \textit{Baetica} province we want to identify links between workshops based on the similarity of their stamps. In the consumption Roman provinces our aim is to detect groups of nearby sites sharing stamps and to what extent stamps are clustered based on provinces. This economic connection could be identified with two ideas : a) nearby sites have more similar amphoric stamps and b) sets of stamps can only be found on specific provinces because they were shipped together.

%In particular, we study the distribution of amphoric stamps to identify a correlation between geographical distance and similarity.

We proposed three working hypotheses based on these ideas: a) we can identify a correlation between spatial distance and the distribution of amphoric stamps, b) stamp sets found in nearby workshops should be more similar and c) there is a low mobility of amphoric stamps from consumption centres to other regions (i.e. stamps always stay in the same region).

This study proposes a robust baseline to explore the distribution of Baetican olive oil production by identifying which centres have more similar stamps and then confront this measure against their proximity. To do this, a population-based approach has been used to analyse the dispersion of stamps between amphora workshops \citep{rubio-campillo_ecology_2018}. This quantitative method provides a way to measure the similarity between centre based on the amphoric stamps found on each of them. This ecological approach is based on three steps: a) to quantify stamp similarity between sites, b) to group sites based on this similarity and c) to find correlation patterns between stamp similarity and proximity. The rationale behind the method is as follows: if trade structure was centrally planned then we should be able to stablish links between workshops and consumption sites based on the similarity of its stamps; if this is not the case then results would suggest that workshops worked independently and agreements were made at an individual level.

%Stamps will be used to identify economic patterns by analysing their similarity.

The paper discusses these questions, the analysis and results as follows: next section introduces the historical context. Section two defines the dataset and the methods used for the analysis. Section three presents the results and section 4 discusses them; the paper finishes with some concluding remarks on the research performed.

@@

The \textit{Baetican} province was one of the most suitable regions to tackle the challenge of the huge demand for olive oil across the Roman provinces. For this reason, the area saw an increase in productivity as a massive specialized infrastructure of olive oil production was gradually deployed. This production of olive oil had the largest peark from the first to the third century AD~\citep{remesal_concierto}.

@@

The chronology of the Baetican workshops is typically defined between the first to the third centuries AD~\citep{millet_anforas_1998,rodriguez_baetican_1998,chic2005comercio}. Archaeological evidence suggests that this highly specialised production was developed across a long activity period\citep{remesal_anforas_2004}, but at the same time the process seemed to have been planned at some level as is observed by the relatively minor changes in amphoric typologies and the high degree of standardisation on the amphorae found across at least 50 workshops.

@@

Dressel 20 was the amphora most stamped during the Roman Empire ~\citep[18]{millet_anforas_1998}. The large sample size of recovered Dressel 20 stamps is distributed across all Western provinces of the Roman Empire. This wide distribution is observed in the large volume of publications discussing the origin, use, chronology and spatial distribution of the stamped codes~\citep{dressel_ricerche_1878,rodriguez_economioleicola_1977,chicepi1985,millet_anforas_1998, remesal_sellar_2016} (Fig.\ref{amphora}).

\begin{figure}[htp]
	\centering
\includegraphics[scale=0.5]{dressel20}
\caption{Dressel 20 were typically marked with stamps of three letters called \textit{tria nomina}}
\label{amphora}
\end{figure}

Most stamps seemed to have been used across a long activity period which can be difficult to date with the required level of accuracy. They have been frequently dated using as reference consular dates found on nearby \textit{tituli picti} recovered from \textit{Monte Testaccio}~\citep{Testaccio1, berni_millet_epigrafianforica_2008}.
The chronological information of \textit{tituli picti} is very useful, but the small amount of dates recovered and the fact that a vast majority of them come from the same site poses some challenges to this precise dating. In addition, the fact that \textit{Monte Testaccio} was a dump suggests that the diposition of amphorae vessels may have been altered at a later phase due to changes in the long-term dynamics of creation of this artificial mound.

@@

The stamp codes are interpreted based on three main ideas: content (olive oil), context (amphora workshop) and subject (individuals involved in the production). On the one hand, it seems that stamps could have been identified as the landowner of the olive groves \citep{rodriguez_economioleicola_1977}. On the other hand, they could also belong to the owner of the workshops where the amphorae were made or even to a group of amphora workers~\citep{berni_millet_epigrafianforica_2008}. Despite the unsolved question of ownership some interesting ideas can be explored based on the finding of similar stamps across the Empire. Observable patterns can be used to explore how this production was organised and whether it is possible to identify specific production processes fdor olive oil trade. Our questions here will be focused on the distribution of amphoric stamps; were different stamps shipped through different distribution routes? Were specific stamp codes used in different workshops?

>>>>

The aim of this study is to explore the trade dynamics involved in the large-scale distribution of olive oil produced in \textit{Baetica} province (currently Andalusia). \textit{Baetica} became an important production and distribution centre during the Roman Empire and its goods have been found in thousands of sites, but it is still unclear how exactly this large scale production was organised.

We explore here connections between production and consumption sites during the period of most intense activity (1st to 3rd centuries AD). Specifically, we use amphoric stamps as proxies to identify the workshops where amphorae located in consumer sites were produced.

To achieve this goal, we analyse a dataset of stamps found in two types of sites: 1) workshops from \textit{Baetica} province and 2) consumption centres located in two provinces \textit{Germania} (Inferior and Superior) and \textit{Britannia}. This dataset is used to detect links between specific sites that may share a large percentage of common stamps. The analysis is performed using quantitative methods borrowed from Ecology that allows us to group sites based on the number of stamps shared between them.

%The analysis explores how the quantitative approach provides a useful tool for the interpretation of the economic processes.

Finally, results highlight the structure of economic processes of the Roman Empire using the archaeological evidence of olive oil large-scale distribution.

@@

The Empire developed a series of structures to support and organise this long-range trade and specialised entire regions to produce specific goods at a massive scale. The province of \textit{Baetica} (currently Andalusia, southern Spain) is an example of this specialisation process. \textit{Baetica} became one of the most important olive oil production centres during the Roman Empire; olive oil was an essential good for Romans because it was used in almost every aspect of their daily life such as cooking, hygiene or lighting~\citep{mattingly_d.j._oil_1988}. This high demand required a huge increase in the production which needed to be distributed to cities across thousands of kilometres. This large-scale trade was organised via the shipment of millions of amphorae using maritime and riverine routes to all the provinces and particularly to Italy and the military garrisons guarding the borders of the Empire~\citep{blazquez_exportacion_1980}.

The structure and processes associated with this massive olive oil production have been extensively discussed over the last decades\citep{rodriguez_economioleicola_1977, duncan1982economy, Chic_hispania_1997,millet_anforas_1998}. However, the complexity and sheer scale of the process makes its understanding challenging, 
%se repite despite aquí dos veces
despite several archaeological excavations and the existence of some written records discussing its structure. Our knowledge of this economic activity has benefited from new findings and data sources, but despite all these advances several questions remain open: How was this massive amphora production organised? Can we detect the economic ties between provinces? Was this distribution centralised or each region organised it differently?

Recent advances in Roman studies have currently led to an environment with various interpretations on these commercial dynamics~\citep{temin_economy_2006,quantifyingwilson2009}. %The application of diverse novel quantitative approaches to new accessible datasets has allowed us to improve our interpretation of the growing amount of archaeological evidence and contribute to a better understanding of the dynamics of Roman production
The application of novel quantitative approaches has allowed us to improve our interpretation of the growing amount of archaeological evidence and contribute to a better understanding of the Roman production
\citep{brughmans_roman_2016,orengo_seeds_2016,bayesian_2018,coto-sarmiento_identifying_2018,rubio-campillo_ecology_2018}. This work aligns with these efforts and focuses on the distribution of olive oil towards Roman provinces in the borders. Amphoric production in \textit{Baetica} province will help us identify links between production centres and consumption centres in order to detect an economic pattern in the commerce by analysing amphoric stamps finds in these provinces.

The finding of identical amphoric stamps in different provinces has allowed us to explore patterns of olive oil distribution within the Empire. Despite this archaeological evidence, there is still no consensus on the structure that generated this pattern; how was the economic relation between production and consumption centres? Did they follow a centralised process to distribute olive oil in those provinces or each workshop organised its own trade networks?
%repite aims dos veces  
This paper aims to explore these questions regarding the olive oil market by identifying links between production centres and consumption provinces using similarity indexes for amphoric stamps. In this way, our work aims to detect regional patterns for a specific commercial product (i.e. olive oil) which was shipped between distant provinces~\citep{isaksen_network_2006}.

@@

In \textit{Baetica} province we want to identify links between workshops based on the similarity of their stamps. In the consumption Roman provinces, our aim is to detect groups of nearby sites sharing stamps and to what extent stamps are clustered based on provinces. This economic connection could be identified with two ideas: a) nearby sites have more similar amphoric stamps and b) sets of stamps can only be found on specific provinces because they were shipped together.
%find in or find on

We proposed three working hypotheses based on these ideas: a) we can identify a correlation between spatial distance and the distribution of amphoric stamps, b) stamp sets found in nearby workshops should be more similar and c) there is low mobility of amphoric stamps from consumption centres to other regions (i.e. stamps always stay in the same region).

This study proposes a robust baseline to explore the distribution of Baetican olive oil production by identifying which centres have more similar stamps and then confront this measure against their proximity. To do this, a population-based approach has been used to analyse the dispersion of stamps between amphora workshops \citep{rubio-campillo_ecology_2018}. This quantitative method provides a way to measure the similarity between centre based on the amphoric stamps found on each of them. This ecological approach is based on three steps: a) to quantify stamp similarity between sites, b) to group sites based on this similarity and c) to find correlation patterns between stamp similarity and proximity. The rationale behind the method is as follows: if trade structure was centrally planned then we should be able to establish links between workshops and consumption sites based on the similarity of its stamps; if this is not the case then results would suggest that workshops worked independently and agreements were made at an individual level.

The paper addresses these questions, the analysis and results as follows: next section introduces the historical context. Section two defines the dataset and the methods used for the analysis. Section three presents the results and section four discusses them; the paper finishes with some concluding remarks on the research performed.

@@

The \textit{Baetican} province was one of the most suitable regions to tackle the challenge of the huge demand for olive oil across the Roman provinces. For this reason, the area saw an increase in productivity as a massive specialised infrastructure of olive oil production was gradually deployed. This production of olive oil had the largest peak from the first to the third century AD~\citep{remesal_concierto}.

@@

The chronology of the Baetican workshops is typically defined between the first to the third centuries AD~\citep{millet_anforas_1998,rodriguez_baetican_1998,chic2005comercio}. Archaeological evidence suggests that this highly specialised production was developed across a long activity period\citep{remesal_anforas_2004}. The process seemed to have been planned at some level as is observed by the relatively minor changes in amphoric typologies and the high degree of standardisation in the production.

@@

Dressel 20 was the amphora most stamped during the Roman Empire~\citep[18]{millet_anforas_1998}. The large sample size of recovered Dressel 20 stamps is distributed across all Western provinces of the Roman Empire. This wide distribution is observed in the large volume of publications discussing the origin, use, chronology and spatial distribution of the stamped codes~\citep{dressel_ricerche_1878,rodriguez_economioleicola_1977,chicepi1985,millet_anforas_1998, remesal_sellar_2016} (Fig.\ref{amphora}).

\begin{figure}[htp]
	\centering
\includegraphics[scale=0.5]{dressel20}
\caption{Dressel 20 were typically marked with stamps of three letters called \textit{tria nomina}}
\label{amphora}
\end{figure} 
%found in find in
Most stamps seemed to have been used across a long activity period which can be difficult to date with the required level of accuracy. They have been frequently dated using as reference consular dates found on nearby \textit{tituli picti} recovered from \textit{Monte Testaccio}~\citep{Testaccio1, berni_millet_epigrafianforica_2008}.
The chronological information of \textit{tituli picti} is very useful, but the small amount of dates recovered and the fact that a vast majority of them come from the same site poses some challenges to this precise dating. In addition, the fact that \textit{Monte Testaccio} was a dump suggests that the disposition of amphorae vessels may have been altered at a later phase due to changes in the long-term dynamics of creation of this artificial mound.

@@

The stamp codes are interpreted based on three main ideas: content (olive oil), context (amphora workshop) and subject (individuals involved in the production). On the one hand, it seems that stamps could have been identified as the landowner of the olive groves \citep{rodriguez_economioleicola_1977}. On the other hand, they could also belong to the owner of the workshops where the amphorae were made or even to a group of amphora workers~\citep{berni_millet_epigrafianforica_2008}. Despite the unsolved question of ownership, some interesting ideas can be explored based on the finding of similar stamps across the Empire. Observable patterns can be used to explore how this production was organised and whether it is possible to identify specific production processes for olive oil trade. Our questions here will be focused on the distribution of amphoric stamps; were different stamps shipped through different distribution routes? Were specific stamp codes used in different workshops?
